\bf A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-05-11 paper/paper-jsdse/paper-jsdse.tex
Finished address all of the comments.

Our approach to cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al. 2019). The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables (Wickham 2011). Tidy data shouldn't be confused with ``tame data,'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called \texttt{yowie}, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

In the context of official statistics, M. van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables). What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain. In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida} and Section \ref{censor}.

@@

Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the \texttt{tidyverse} suite of packages (Wickham et al. 2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham et al. 2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{https://github.com/numbats/yowie/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

Similarly, there are issues for the \texttt{race} variable. The current US Federal guidelines (Office of Management and Budget 1997) state that there should be at least five categories for race, and an individual should be able to identify as more than one race category. in addition, ``Hispanic'' or ``Latino'' should be considered to be an ethnic group rather than a race. This level of detail is not available in the database. There is one variable with only three categories, and an individual can be a member of only one. Thus, the \texttt{race} variable is inadequate by today's standards, and educators should point to the current Federal guidelines when using this variable, and keep in mind that the purpose of any analysis is to study racism rather than race.

@@

The next step is processing the highest grade completed. There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis. The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively. For each individual, there is only one value of \texttt{hgc} and \texttt{hgc\_i}. This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on). We decide to include both

The second definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as \texttt{grade}). This value can change over time, but it should only increase. This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals. This is recorded in \texttt{new\_data\_qnames} as columns beginning with \texttt{Q3-4} and \texttt{HGC} with year as a suffix. In addition, it is also in the columns beginning with \texttt{HGCREV} reflecting revised data. We chose to use these revised values because there were fewer missing values indicating it had been more thoroughly checked. In the survey years when the revised value were not reported, i.e., in 2012, 2014, 2016, and 2018, we used the unrevised version accordingly.

@@

\caption{\label{tab:age-table}Frequency table of the age at the start of the survey in NSLY79 cohort}
\centering
\begin{tabular}[t]{rr}
\toprule
Age & Number of individuals\\
\midrule
15 & 1,265\\
16 & 1,550\\
17 & 1,600\\
18 & 1,530\\
19 & 1,662\\
20 & 1,722\\
21 & 1,677\\
22 & 1,680\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:gender-race-table}Contingency table for sex and race for the full NLSY79 data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Race} & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){2-4}
Sex & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
\midrule
F & 1,561 (25\%) & 1,002 (16\%) & 3,720 (59\%) & 6,283\\
M & 1,613 (25\%) & 1,000 (16\%) & 3,790 (59\%) & 6,403\\
\midrule
Total & 3,174 (25\%) & 2,002 (16\%) & 7,510 (59\%) & 12,686\\
\bottomrule
\end{tabular}
\end{table}

@@

{\centering \includegraphics[width=0.85\linewidth]{../figures/flowchart_blind}

@@

\caption{The stages of data cleaning from the raw data to get three datasets contained in \texttt{yowie}. ``\# of individuals'' means the number of respondents included in each stage, while ``\# of observations'' means the number of rows in the data. The color represents the stage of data cleaning in the statistical value chain (M. P. J. van der Loo and de Jonge 2021). Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart}
\end{figure}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-iris-data}{}}%
Anderson, Edgar. 1935. {``The Irises of the Gaspe Peninsula.''} \emph{Bulletin of the American Iris Society} 59: 2--5.

\leavevmode\vadjust pre{\hypertarget{ref-nlsy79}{}}%
Bureau of Labor Statistics, U.S. Department of Labor. 2021a. {``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University. Columbus, OH, through \url{https://www.nlsinfo.org/bibliography-citing-nls-data}.

\leavevmode\vadjust pre{\hypertarget{ref-nlsy79guide}{}}%
---------. 2021b. {``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.''} \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.

\leavevmode\vadjust pre{\hypertarget{ref-shiny}{}}%
Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{{shiny: Web Application Framework for R}}. \url{https://CRAN.R-project.org/package=shiny}.

\leavevmode\vadjust pre{\hypertarget{ref-Chatfield1985TIEo}{}}%
Chatfield, C. 1985. {``The Initial Examination of Data.''} \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.

\leavevmode\vadjust pre{\hypertarget{ref-eliznlsy}{}}%
Cooksey, Elizabeth C. 2017. {``Using the National Longitudinal Surveys of Youth ({NLSY}) to Conduct Life Course Analyses.''} In \emph{Handbook of Life Course Health Development}, edited by Richard M. Lerner Neal Halfon Christoper B. Forrest, 561--77. Cham: Springer. https://doi.org/\url{https://doi.org/10.1007/978-3-319-47143-3_23}.

\leavevmode\vadjust pre{\hypertarget{ref-DasuTamraparni2003Edma}{}}%
Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.

\leavevmode\vadjust pre{\hypertarget{ref-janitor}{}}%
Firke, Sam. 2020. \emph{{janitor: Simple Tools for Examining and Cleaning Dirty Data}}. \url{https://CRAN.R-project.org/package=janitor}.

\leavevmode\vadjust pre{\hypertarget{ref-racismnotrace}{}}%
Fullilove, M. T. 1998. {``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.''} \emph{American Journal of Public Health} 88 (9): 1297--98.

\leavevmode\vadjust pre{\hypertarget{ref-grimshaw}{}}%
Grimshaw, Scott D. 2015. {``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14. https://doi.org/\url{https://doi.org/10.1080/00031305.2015.1081106}.

\leavevmode\vadjust pre{\hypertarget{ref-SAGER}{}}%
Heidari, S., T. F. Babor, P. De Castro, S. Tort, and M. Curno. 2016. {``Sex and Gender Equity in Research: Rationale for the {SAGER} Guidelines and Recommended Use.''} \emph{Research Integrity and Peer Review} 1 (2). https://doi.org/\url{https://doi.org/10.1186/s41073-016-0007-6}.

\leavevmode\vadjust pre{\hypertarget{ref-purrr}{}}%
Henry, Lionel, and Hadley Wickham. 2020. \emph{{purrr: Functional Programming Tools}}. \url{https://CRAN.R-project.org/package=purrr}.

\leavevmode\vadjust pre{\hypertarget{ref-penguins-data}{}}%
Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.

\leavevmode\vadjust pre{\hypertarget{ref-HuebnerMariannePhD2016Asat}{}}%
Huebner, Marianne, Werner Vach, and Saskia le Cessie. 2016. {``A Systematic Approach to Initial Data Analysis Is Good Research Practice.''} \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.

\leavevmode\vadjust pre{\hypertarget{ref-ilk2004}{}}%
Ilk, Ozlem. 2004. {``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.''} PhD thesis, Iowa State University. \url{https://doi.org/10.31274/rtd-180813-11012}.

\leavevmode\vadjust pre{\hypertarget{ref-kennedy2020using}{}}%
Kennedy, Lauren, Katharine Khanna, Daniel Simpson, and Andrew Gelman. 2020. {``Using Sex and Gender in Survey Adjustment.''} \url{https://arxiv.org/abs/2009.14401}.

\leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%
Kim, A. Y, C. Ismay, and J. Chunn. 2018. {``The Fivethirtyeight {R} Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.''} \emph{Technology Innovations in Statistics Education} 11 (1). https://doi.org/\url{https://doi.org/10.5070/T511103589}.

\leavevmode\vadjust pre{\hypertarget{ref-KollerManuel2016rARP}{}}%
Koller, Manuel. 2016. {``{robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models}.''} \emph{Journal of Statistical Software} 75 (6): 1--24.

\leavevmode\vadjust pre{\hypertarget{ref-notaverage}{}}%
Moncrief, Marc. 2015. {``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.''} \url{https://bit.ly/smh-not-normal}.

\leavevmode\vadjust pre{\hypertarget{ref-doh}{}}%
Office of Management and Budget. 1997. {``Revisions to the Standards for the Classification of Federal Data on Race and Ethnicity.''} \url{https://www.govinfo.gov/content/pkg/FR-1997-10-30/pdf/97-28653.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-opendata}{}}%
Open Knowledge Foundation. 2021. {``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.''} 2021. \url{http://opendefinition.org/od/2.1/en/}.

\leavevmode\vadjust pre{\hypertarget{ref-patchwork}{}}%
Pedersen, Thomas Lin. 2020. \emph{{patchwork: The Composer of Plots}}. \url{https://CRAN.R-project.org/package=patchwork}.

\leavevmode\vadjust pre{\hypertarget{ref-MichaelRPergamit2001DWTN}{}}%
Pergamit, Michael R., Charles R. Pierret, Donna S. Rothstein, and Jonathan R. Veum. 2001. {``Data Watch: The National Longitudinal Surveys.''} \emph{The Journal of Economic Perspectives} 15 (2): 239--53.

\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.

\leavevmode\vadjust pre{\hypertarget{ref-SingerJudithD2003Alda}{}}%
Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.

\leavevmode\vadjust pre{\hypertarget{ref-notiris}{}}%
Stodel, Megan. 2020. {``Stop Using Iris.''} \url{https://www.meganstodel.com/posts/no-to-iris/}.

\leavevmode\vadjust pre{\hypertarget{ref-brolgar}{}}%
Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}}. \url{https://github.com/njtierney/brolgar}.

\leavevmode\vadjust pre{\hypertarget{ref-tukey}{}}%
Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.

\leavevmode\vadjust pre{\hypertarget{ref-validate}{}}%
van der Loo, Mark P. J., and Edwin de Jonge. 2021. {``Data Validation Infrastructure for {R}.''} \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.

\leavevmode\vadjust pre{\hypertarget{ref-LooMarkvander2018Sdcw}{}}%
van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in r}.

\leavevmode\vadjust pre{\hypertarget{ref-mass}{}}%
Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with s}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.

\leavevmode\vadjust pre{\hypertarget{ref-tsibble}{}}%
Wang, Earo, Dianne Cook, and Rob J Hyndman. 2020. {``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.''} \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78. \url{https://doi.org/10.1080/10618600.2019.1695624}.

\leavevmode\vadjust pre{\hypertarget{ref-plyr}{}}%
Wickham, Hadley. 2011. {``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. https://doi.org/\url{https://doi.org/10.18637/jss.v040.i01}.

\leavevmode\vadjust pre{\hypertarget{ref-WickhamHadley2014TD}{}}%
---------. 2014. {``Tidy Data.''} \emph{Journal of Statistical Software} 59 (10): 1--23.

\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
---------. 2016. \emph{{ggplot2: Elegant Graphics for Data Analysis}}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.

\leavevmode\vadjust pre{\hypertarget{ref-stringr}{}}%
---------. 2019. \emph{{stringr: Simple, Consistent Wrappers for Common String Operations}}. \url{https://CRAN.R-project.org/package=stringr}.

\leavevmode\vadjust pre{\hypertarget{ref-tidyr}{}}%
---------. 2020. \emph{{tidyr: Tidy Messy Data}}. \url{https://CRAN.R-project.org/package=tidyr}.

\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019. {``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software} 4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.

\leavevmode\vadjust pre{\hypertarget{ref-dplyr}{}}%
Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{{dplyr: A Grammar of Data Manipulation}}. \url{https://CRAN.R-project.org/package=dplyr}.

\leavevmode\vadjust pre{\hypertarget{ref-readr}{}}%
Wickham, Hadley, and Jim Hester. 2020. \emph{{readr: Read Rectangular Text Data}}. \url{https://CRAN.R-project.org/package=readr}.

>>>>

Our approach to cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al. 2019). The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables (Wickham 2011). Tidy data shouldn't be confused with ``tame data,'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called \texttt{yowie}, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

In the context of official statistics, van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables). What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain. In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida} and Section \ref{censor}.

@@

Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al. 2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham, François, et al. 2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{https://github.com/numbats/yowie/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

Similarly, there are issues for the \texttt{race} variable. The current US Federal guidelines (Office of Management and Budget 1997) state that there should be at least five categories for race, and an individual should be able to identify as more than one race category. In addition, ``Hispanic'' or ``Latino'' should be considered to be an ethnic group rather than a race. This level of detail is not available in the database. There is one variable with only three categories, and an individual can be a member of only one. Thus, the \texttt{race} variable is inadequate by today's standards, and educators should point to the current Federal guidelines when using this variable, and keep in mind that the purpose of any analysis is to study racism rather than race.

@@

The next step is processing the highest grade completed. There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis. The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively. For each individual, there is only one value of \texttt{hgc} and \texttt{hgc\_i}. This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on). We decide to include both the factor and integer values to be used as examples in various classroom demonstrations, such as to explain different data types. Besides, it can also be used as the demonstration of different type of data visualization which is determined by the data type. It can also be used for teaching the modeling of cross sectional data.

The second definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as \texttt{grade}). This value can change over time, but it should only increase. This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals. Hence, different from \texttt{hgc}, this variable could be used to demonstrate longitudinal and time series analysis in the classroom. This is recorded in \texttt{new\_data\_qnames} as columns beginning with \texttt{Q3-4} and \texttt{HGC} with year as a suffix. In addition, it is also in the columns beginning with \texttt{HGCREV} reflecting revised data. We chose to use these revised values because there were fewer missing values indicating it had been more thoroughly checked. In the survey years when the revised value were not reported, i.e., in 2012, 2014, 2016, and 2018, we used the unrevised version accordingly.

@@

\caption{\label{tab:age-table}Frequency table of the age at the start of the survey in NSLY79 cohort in the extracted data.}
\centering
\begin{tabular}[t]{rr}
\toprule
Age & Number of individuals\\
\midrule
15 & 1,265\\
16 & 1,550\\
17 & 1,600\\
18 & 1,530\\
19 & 1,662\\
20 & 1,722\\
21 & 1,677\\
22 & 1,680\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:gender-race-table}Contingency table for sex and race for the extracted NLSY79 demographic data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Race} & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){2-4}
Sex & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
\midrule
F & 1,561 (25\%) & 1,002 (16\%) & 3,720 (59\%) & 6,283\\
M & 1,613 (25\%) & 1,000 (16\%) & 3,790 (59\%) & 6,403\\
\midrule
Total & 3,174 (25\%) & 2,002 (16\%) & 7,510 (59\%) & 12,686\\
\bottomrule
\end{tabular}
\end{table}

@@

{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie-paper/paper/figures/flowchart}

@@

\caption{The stages of data cleaning from the raw data to get three datasets contained in \texttt{yowie}. ``\# of individuals'' means the number of respondents included in each stage, while ``\# of observations'' means the number of rows in the data. The color represents the stage of data cleaning in the statistical value chain (van der Loo and de Jonge 2021). Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart}
\end{figure}

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-iris-data}{}%
Anderson, Edgar. 1935. ``The Irises of the Gaspe Peninsula.'' \emph{Bulletin of the American Iris Society} 59: 2--5.

\leavevmode\hypertarget{ref-nlsy79}{}%
Bureau of Labor Statistics, U.S. Department of Labor. 2021a. ``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University. Columbus, OH, through \url{https://www.nlsinfo.org/bibliography-citing-nls-data}.

\leavevmode\hypertarget{ref-nlsy79guide}{}%
---------. 2021b. ``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.'' \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.

\leavevmode\hypertarget{ref-shiny}{}%
Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.

\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.

\leavevmode\hypertarget{ref-eliznlsy}{}%
Cooksey, Elizabeth C. 2017. ``Using the National Longitudinal Surveys of Youth (NLSY) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M. Lerner Neal Halfon Christoper B. Forrest, 561--77. Cham: Springer. \url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.

\leavevmode\hypertarget{ref-DasuTamraparni2003Edma}{}%
Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.

\leavevmode\hypertarget{ref-janitor}{}%
Firke, Sam. 2020. \emph{janitor: Simple Tools for Examining and Cleaning Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.

\leavevmode\hypertarget{ref-racismnotrace}{}%
Fullilove, M. T. 1998. ``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.

\leavevmode\hypertarget{ref-grimshaw}{}%
Grimshaw, Scott D. 2015. ``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14. \url{https://doi.org/https://doi.org/10.1080/00031305.2015.1081106}.

\leavevmode\hypertarget{ref-SAGER}{}%
Heidari, S., T. F. Babor, P. De Castro, S. Tort, and M. Curno. 2016. ``Sex and Gender Equity in Research: Rationale for the SAGER Guidelines and Recommended Use.'' \emph{Research Integrity and Peer Review} 1 (2). \url{https://doi.org/https://doi.org/10.1186/s41073-016-0007-6}.

\leavevmode\hypertarget{ref-purrr}{}%
Henry, Lionel, and Hadley Wickham. 2020. \emph{purrr: Functional Programming Tools}. \url{https://CRAN.R-project.org/package=purrr}.

\leavevmode\hypertarget{ref-penguins-data}{}%
Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.

\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
Huebner, Marianne, Werner Vach, and Saskia le Cessie. 2016. ``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.

\leavevmode\hypertarget{ref-ilk2004}{}%
Ilk, Ozlem. 2004. ``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.'' PhD thesis, Iowa State University. \url{https://doi.org/10.31274/rtd-180813-11012}.

\leavevmode\hypertarget{ref-kennedy2020using}{}%
Kennedy, Lauren, Katharine Khanna, Daniel Simpson, and Andrew Gelman. 2020. ``Using Sex and Gender in Survey Adjustment.'' \url{http://arxiv.org/abs/2009.14401}.

\leavevmode\hypertarget{ref-tamedata}{}%
Kim, A. Y, C. Ismay, and J. Chunn. 2018. ``The Fivethirtyeight R Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.'' \emph{Technology Innovations in Statistics Education} 11 (1). \url{https://doi.org/https://doi.org/10.5070/T511103589}.

\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
Koller, Manuel. 2016. ``robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.

\leavevmode\hypertarget{ref-notaverage}{}%
Moncrief, Marc. 2015. ``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.'' \url{https://bit.ly/smh-not-normal}.

\leavevmode\hypertarget{ref-doh}{}%
Office of Management and Budget. 1997. ``Revisions to the Standards for the Classification of Federal Data on Race and Ethnicity.'' \url{https://www.govinfo.gov/content/pkg/FR-1997-10-30/pdf/97-28653.pdf}.

\leavevmode\hypertarget{ref-opendata}{}%
Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.

\leavevmode\hypertarget{ref-patchwork}{}%
Pedersen, Thomas Lin. 2020. \emph{patchwork: The Composer of Plots}. \url{https://CRAN.R-project.org/package=patchwork}.

\leavevmode\hypertarget{ref-MichaelRPergamit2001DWTN}{}%
Pergamit, Michael R., Charles R. Pierret, Donna S. Rothstein, and Jonathan R. Veum. 2001. ``Data Watch: The National Longitudinal Surveys.'' \emph{The Journal of Economic Perspectives} 15 (2): 239--53.

\leavevmode\hypertarget{ref-R}{}%
R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.

\leavevmode\hypertarget{ref-SingerJudithD2003Alda}{}%
Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.

\leavevmode\hypertarget{ref-notiris}{}%
Stodel, Megan. 2020. ``Stop Using Iris.'' \url{https://www.meganstodel.com/posts/no-to-iris/}.

\leavevmode\hypertarget{ref-brolgar}{}%
Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}. \url{https://github.com/njtierney/brolgar}.

\leavevmode\hypertarget{ref-tukey}{}%
Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.

\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.

\leavevmode\hypertarget{ref-validate}{}%
van der Loo, Mark P. J., and Edwin de Jonge. 2021. ``Data Validation Infrastructure for R.'' \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.

\leavevmode\hypertarget{ref-mass}{}%
Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.

\leavevmode\hypertarget{ref-tsibble}{}%
Wang, Earo, Dianne Cook, and Rob J Hyndman. 2020. ``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.'' \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78. \url{https://doi.org/10.1080/10618600.2019.1695624}.

\leavevmode\hypertarget{ref-plyr}{}%
Wickham, Hadley. 2011. ``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/https://doi.org/10.18637/jss.v040.i01}.

\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
---------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software} 59 (10): 1--23.

\leavevmode\hypertarget{ref-ggplot2}{}%
---------. 2016. \emph{ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.

\leavevmode\hypertarget{ref-stringr}{}%
---------. 2019. \emph{stringr: Simple, Consistent Wrappers for Common String Operations}. \url{https://CRAN.R-project.org/package=stringr}.

\leavevmode\hypertarget{ref-tidyr}{}%
---------. 2020. \emph{tidyr: Tidy Messy Data}. \url{https://CRAN.R-project.org/package=tidyr}.

\leavevmode\hypertarget{ref-tidyverse}{}%
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019. ``Welcome to the tidyverse.'' \emph{Journal of Open Source Software} 4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.

\leavevmode\hypertarget{ref-dplyr}{}%
Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{dplyr: A Grammar of Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.

\leavevmode\hypertarget{ref-readr}{}%
Wickham, Hadley, and Jim Hester. 2020. \emph{readr: Read Rectangular Text Data}. \url{https://CRAN.R-project.org/package=readr}.

\leavevmode\hypertarget{ref-nlsy79edu}{}%
Wolpin, Kenneth I. 2005. ``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Published by Bureau of Labor Statistics, U.S. Department of Labor. \url{https://www.bls.gov/opub/mlr/2005/02/art3full.pdf}.

\leavevmode\hypertarget{ref-knitr}{}%
Xie, Yihui. 2014. ``Knitr: A Comprehensive Tool for Reproducible Research in R.'' In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D. Peng. Chapman; Hall/CRC. \url{http://www.crcpress.com/product/isbn/9781466561595}.

\leavevmode\hypertarget{ref-rmarkdown}{}%
Xie, Yihui, Christophe Dervieux, and Emily Riederer. 2020. \emph{R Markdown Cookbook}. Boca Raton, Florida: Chapman; Hall/CRC. \url{https://bookdown.org/yihui/rmarkdown-cookbook}.
